### 2022-02-23 ehoh/ehoh.tex
more

>>>>

Saturating provers show the unsatisfiability of a clause set
by systematically adding logical consequences,
% (up to simplification and redundancy),
eventually deriving the empty clause
as a witness of unsatisfiability. They implement two kinds of
inference rules: \emph{Generating rules} produce new clauses and are
needed for completeness, whereas \emph{simplification rules} delete
existing clauses or replace them by simpler clauses. This
simplification is crucial for success, and most modern provers spend a
large part of their time on simplification. E's main loop, which applies the
rules, implements the \relax{given clause} procedure, as described in
Sect.~\ref{sec:pre:saturation}.

% The proof state is represented by two disjoint
% subsets of clauses, the set of \emph{processed} clauses $P$ and the set of
% \emph{unprocessed} clauses $U$. Initially, all clauses are unprocessed. At each
% iteration of the loop, the prover heuristically selects a \emph{given clause}
% from $U$, adds it to $P$, and performs all generating inferences between this
% clause and all clauses in $P$. Resulting new clauses are added to $U$. This
% maintains the invariant that all direct consequences between clauses in $P$ have
% been performed. Simplification is performed on the given clause (using clauses
% in $P$ as side premises), on clauses in $P$ (using the given clause), and on
% newly generated clauses (again, using $P$).

Ehoh is based on the same logical calculus as E, except that it is
generalized to \lfhol{} terms. The standard inference rules and completeness
proof of superposition with respect to intensional Boolean-free \lfhol{} fragment of our
logic can be reused verbatim; the only changes concern the basic definitions of
terms and substitutions \cite[Sect.~1]{bbcw-21-lfho}.
\begin{rep}Refutational completeness of superposition for \lfhol{} terms has
been formally proved by Peltier \cite{np-2016-sup-formalisation} using Isabelle.\end{rep}
%
We introduced support for first-class Boolean terms in Ehoh by extending the
preprocessor, as explained in Sect.~\ref{sec:ehoh:preprocessing}.

\ourpara{The Generating Rules}

The superposition rules were introduced in Sect.~\ref{sec:pre:rules}. For completeness,
we repeat them with slightly simplified notation, as we do not repeat side conditions:

\hfill
\begin{tabular}{ll}
  \namedinference{SN}
  {s \eq t \mathrel\vee C \qquad u[s'] \noteq v \mathrel\vee D}
  {\sigma(u[t] \mathbin{\noteq} v \mathrel\vee C \mathrel\vee D)}
  &
  \namedinference{ER}{s \noteq s' \mathrel\vee C}{\sigma(C)} \\[2\jot]
  \namedinference{SP}
    {s \eq t \mathrel\vee C \qquad u[s'] \eq v \mathrel\vee D}
    {\sigma(u[t] \mathbin{\eq} v \mathrel\vee C \mathrel\vee D)}
  &
  \namedinference{EF}{s \eq t \mathrel\vee  s' \eq u \mathrel\vee C}{\sigma(t \noteq u \mathrel\vee  s \eq u \mathrel\vee C)}
\end{tabular}
\hfill
\kern\belowdisplayskip

\noindent
In each rule, $\sigma$ denotes the MGU of $s$ and $s'$.

Equality resolution (ER) and equality factoring (EF) \begin{rep}are single-premise rules
that \end{rep}work on the entire left- or right-hand side of a literal of
the given clause. To generalize them, it suffices to disable prefix
optimization for unification.
\confrep{By contrast, t}{\par T}%
he rules for superposition into negative and positive literals (SN~and~SP) are more
complex. As two-premise rules, they require the prover to find a partner
for the given clause. There are two cases to consider\confrep{.}{, depending on whether
the given clause acts as the first or second premise in an inference.
Moreover, since the rules operate on subterms~$s'$ of a clause, the prover
must be able to efficiently locate all relevant subterms, including
\lfhol{} prefix subterms.}
%
To cover the case where the given clause acts as the left premise, the prover
relies on a fingerprint index to compute a set of clauses containing terms
possibly unifiable with a side $s$ of a positive literal of the given clause.
Thanks to our generalization of fingerprints, in Ehoh this candidate set is
guaranteed to overapproximate the set of all possible inference partners. The
unification algorithm is then applied to filter out unsuitable candidates.
Thanks to prefix optimization, we can avoid polluting the
index with all prefix subterms.

When the given clause is the right premise, the prover traverses
its subterms $s'$ looking for inference partners in another fingerprint index,
which contains only entire left-~and right-hand sides of equalities. Like E,
Ehoh traverses subterms in a first-order fashion. If prefix unification
succeeds, Ehoh determines the unified prefix and applies the appropriate
inference instance.

\ourpara{The Simplifying Rules}

Unlike generating rules, simplifying rules do not necessarily add
conclusions to the proof state---they can also remove premises. E
implements over a dozen simplifying rules, with unconditional
rewriting and clause subsumption as the most significant
examples. Here, we restrict our attention to a single rule, which best illustrates
the challenges of supporting \lfhol:
%
\[\namedsimp{ES}{s \eq t \qquad u[\sigma(s)] \eq u[\sigma(t)] \mathrel\vee C}{s \eq t}\]
%
Given an equation $s \eq t$, equality subsumption (\infname{ES}) removes a clause
containing a literal whose two sides are equal except that an instance of $s$
appears on one side where the corresponding instance of $t$ appears on the
other side.

E maintains a perfect discrimination tree storing clauses of the form $s
\eq t$ indexed by $s$ and $t$.
%Each node corresponding to an indexed term
%contains the pointer to unit clause it appears in.
When applying \infname{ES},
E considers each \begin{rep}positive \end{rep}literal $u \eq v$ of the given
clause in turn. It starts by taking the left-hand side~$u$ as a query term.
If an equation $s \eq t$ (or $t \eq s$) is found in the tree, with $\sigma(s)
= u$, the prover checks whether $\sigma'(t) = v$ for some \confrep{}{(possibly
nonstrict) }extension $\sigma'$ of~$\sigma$. If so, ES is
applicable\confrep{}{, with a second premise of the form $\sigma(s) \eq
\sigma(t) \mathrel\vee C$}.
\begin{rep}\par\end{rep}
To consider nonempty contexts, the prover traverses the subterms $u'$ and $v'$ of
$u$ and $v$ in lockstep, as long as they appear under identical
contexts. Thanks to prefix optimization, when Ehoh is given a subterm $u'$,
it can find an equation $s \eq t$ in the tree such that
$\sigma(s)$ is equal to some prefix of $u'$, with some arguments
$\overline{u}$ remaining as unmatched. Checking for equality subsumption
%
then amounts to checking that $v' = \sigma'(t) \; \overline{u}$,
for some extension $\sigma'$ of $\sigma$.

For example, let
$\cst{f} \; (\cst{g} \; \cst{a} \; \cst{b}) \eq \cst{f} \; (\cst{h} \; \cst{g} \; \cst{b})$
be the given clause,
and suppose that $x \; \cst{a} \eq \cst{h} \; x$ is indexed.
Under context $\cst{f}\>[\phantom{.}]$, Ehoh considers the subterms
$\cst{g} \; \cst{a} \; \cst{b}$ and $\cst{h} \; x \; \cst{b}$. It finds the
prefix $\cst{g} \; \cst{a}$ of $\cst{g} \; \cst{a} \; \cst{b}$ in the
%
tree, with $\sigma = \{x \mapsto\nobreak \cst{g}\}$.
The prefix $\cst{h} \; \cst{g}$ of $\cst{h} \; \cst{g} \; \cst{b}$ matches the
indexed equation's right-hand side $\cst{h} \; x$ using the same substitution,
and the remaining argument in both subterms,~$\cst{b}$, is identical.
\begin{rep}Ehoh concludes that the given clause is redundant.\end{rep}

\ourpara{Pragmatic Extensions}

Since Ehoh is based on a mono\-morphic logic, the only way to support
extensionality without changing the calculus is to add a set of extensionality
axioms for every function type occurring in the problem
\cite[Sect.~3.1]{bbcw-21-lfho}. The evaluation by Bentkamp et al.\ of such
an approach was discouraging \cite[Sect.~6]{bbcw-21-lfho}, so we decided to
support extensionality via inference rules in Ehoh. We implemented two well-known
incomplete rules we had experimented with in the context of Zipperposition.

The negative and positive extensionality (\infname{NE} and \infname{PE}) rules
are defined as

\vskip\abovedisplayskip
\noindent\hfill
  \namedinference{NE}{s \noteq t \mathrel\vee C}{s \; (\cst{sk} \; \overline{x}) \noteq t \; (\cst{sk} \; \overline{x}) \mathrel\vee C}%
\qquad
  \namedinference{PE}{s \; x \eq t \; x \mathrel\vee C}{s \eq t \mathrel\vee C}%
\hfill\hbox{}
\vskip\belowdisplayskip

\noindent
For \infname{NE}, $\overline{x}$ contains all the variables occurring in $s$ and
$t$, the terms $s$ and $t$ are of function type, $\cst{sk}$ is a
fresh Skolem symbol, % of appropriate type,
and the literal $s \noteq t$ is eligible for
resolution \cite[Sect.~5]{bbtvw-21-sup-lam}. For \infname{PE}, variable $x$
does not occur in any of the $s$, $t$, or $C$, no literals are selected in $C$, and
$s \; x \eq t \; x$ is a maximal literal.

\begin{sloppypar}
Finally, we introduced an injectivity recognition (IR) rule, which detects
injectivity axioms and asserts the existence of the inverse function for
injective function symbols:
\end{sloppypar}

\vskip\abovedisplayskip

\noindent\hbox{}\hfill
\namedinference{IR}{\cst{f} \; \tuple{x}{n} \noteq \cst{f} \; \tuple{y}{n} \mathrel\vee x_i \eq y_i} {\cst{sk} \; (\cst{f} \;\tuple{x}{n}) \; \overline{x}_J \, \eq\, x_i}\hfill\hbox{}

\vskip\belowdisplayskip

\noindent
where $\textsf{sk}$ is a fresh Skolem symbol,
$J$ is the largest subset of $\{1, \ldots, n$\} such that $x_j = y_j$ for every $j \in J$.
We denote the subsequence of $\tuple{x}{n}$ indexed by $J$ by $\tuple{x}{J}$.
Moreover, we require that
$x_i \neq y_i$, all variables in $\tuple{x}{K} \cdot \tuple{y}{K}$ are distinct,
where $K = \{1, \ldots, n\} \setminus J$,
and neither $\tuple{x}{K}$ nor $\tuple{y}{K}$ shares variables with $\tuple{x}{J}$.
For example, given $\cst{add} \; a \; b \noteq \cst{add} \; a \; b' \mathrel\lor a \eq b'$,
\infname{IR} can derive the existence of the inverse
$\cst{sk}_1$ characterized by $\cst{sk}_{1\!} \; (\cst{add} \; a \; b) \; a \eq b$.
%As a second example, from the clause
%$\cst{cons} \; x \; \mathit{xs} \noteq \cst{cons} \; y \; \mathit{ys}
%\mathrel\lor x \eq y$, we can derive
%$\cst{sk}_2 \; (\cst{cons} \; x \; \mathit{xs}) \eq x$.

### 2022-02-23 pre/pre.tex
unif rules
more

A {\em position} in a term is a tuple of natural numbers, with $\varepsilon$ denoting
the empty tuple. {\em Subterms} and positions are inductively defined as follows. Term
$t$ is a subterm of itself at position $\varepsilon$. If $s$ is a subterm of
$t_i$ at position $p$, then $s$ is a subterm of $\cst{f}(\tuplen{t})$ at
position $i.p$. A {\em context} is a term with zero or more subterms replaced by a
hole $\square$. We write $C[\overline{u}_n]$ for the term resulting from filling
in the holes of a context $C$ with the terms $\overline{u}_n$, left to right.
We say a term is {\em ground} if it has no variables.

@@

Applying substitution to a literal is reduced to applying it to both sides of
the (dis)equa\-tion and it is extended pointwise to clauses. It is written in
postfix notation as $\sigmacl{l}$ and $\sigmacl{C}$. We say $\sigmacl{C}$ is a
ground instance of $C$ if $\sigma$ is a grounding substitution.

@@

\subsection{Unification}

Calculi from the resolution family, including superposition, perform inferences
only on unifiable terms. We say terms $s$ and $t$ are {\em unifiable} if there is a
substitution $\sigma$ such that $\sigmaterm{s} = \sigmaterm{t}$; we further say
$\sigma$ is the unifier. A \emph{unification constraint} $s \unif t$ is an
unordered pair of two terms of the same type. A \emph{most general unifier} is a
unifier $\sigma$, such that for any other unifier $\theta$, there is a
substitution $\varrho$ such that $\theta = \varrho\sigma$. When there is such a
$\varrho$, we say $\sigma$ is more general than $\theta$. In first-order logic,
the most general unifier is unique up to variable renaming. Furthermore,
first-order logic admits efficient algorithms for computing the most general
unifier \cite{hv-09-unifalgs}.

In higher-order logic, unification is performed modulo rules of
$\alpha\beta\eta$ conversion. Under these conditions, the uniqueness of the most
general unifier is not guaranteed. Consider the constraint $X \,
(\cst{f} \, \cst{a}) \unif \cst{f} \, (X \, \cst{a})$. Any substitution of the
form $\{ X \mapsto \lamx{\cst{f}^i \, x} \}$ is a unifier, with $\cst{f}^i \, x$
denoting iterated, $i$-fold application of $\cst{f}$. However, neither of these substitutions
is more general than the other. To generalize most general
unifiers to higher-order logic, the concept of tje {\em complete sets of unifiers} (CSU) was
introduced.

@@

\subsection{Inference Rules}
\newcommand{\mgu}{\ensuremath{\mathit{mgu}}}

@@

\begin{tabular}{m{\dimexpr 0.36\linewidth-2\tabcolsep}m{\dimexpr 0.57\linewidth-2\tabcolsep}}
    \multicolumn{2}{l}{{\bf Equality resolution} (ER)} \\[\jot]
    $\namedinference{ER}{s \noteq t \llor C}{\sigmacl{C}}$ & where $\sigma = \mgu(s,t)$, and $(s \noteq t)\sigma$ is eligible for resolution
\end{tabular}

@@

\begin{tabular}{m{\dimexpr 0.36\linewidth-2\tabcolsep}m{\dimexpr 0.57\linewidth-2\tabcolsep}}
    \multicolumn{2}{l}{{\bf Equality factoring} (EF)} \\[\jot]
    $\namedinference{EF}{s \eq t \llor u \eq v \llor C}{\sigmacl{(t \noteq v \llor u \eq v \llor C)}}$ 
        & where $\sigma = \mgu(s,u), \sigma(t) \not\succ \sigma(s)$ and 
        $(s \eq t)\sigma$ is eligible for paramodulation
\end{tabular}

@@

\begin{tabular}{m{\dimexpr 0.36\linewidth-2\tabcolsep}m{\dimexpr 0.57\linewidth-2\tabcolsep}}
    \multicolumn{2}{l}{{\bf Superposition into positive literals} (SP)} \\[\jot]
    $\namedinference{SP}{s \eq t \llor C \qquad u \eq v \llor D}{(u[p \leftarrow t] \eq v \llor C \llor D)\sigma}$ 
        & where $\sigma = \mgu(u|_p, s), \sigma(t) \not\succ \sigma(s), \sigma(v) \not\succ \sigma(u)$,
        $(s \eq t)\sigma$ is eligible for paramodulation, just like $(u \eq v)\sigma$ and $u|_p$ is not a variable
\end{tabular}

@@

\begin{tabular}{m{\dimexpr 0.36\linewidth-2\tabcolsep}m{\dimexpr 0.57\linewidth-2\tabcolsep}}
    \multicolumn{2}{l}{{\bf Superposition into negative literals} (SN)} \\[\jot]
    $\namedinference{SN}{s \eq t \llor C \qquad u \noteq v \llor D}{(u[p \leftarrow t] \noteq v \llor C \llor D)\sigma}$ 
        & where $\sigma = \mgu(u|_p, s), \sigma(t) \not\succ \sigma(s), \sigma(v) \not\succ \sigma(u)$,
         $(s \eq t)\sigma$ is eligible for paramodulation, $(u \noteq v)\sigma$ is eligible for resolution, and $u|_p$ is not a variable
\end{tabular}

@@

\[
\namedsimp{RN}{s \eq t \qquad u \noteq v \llor D}
              {s \eq t \qquad (u[p \leftarrow \sigma(t)] \eq v \llor C \llor D)\sigma}
\]
where $\sigma = \mgu(u|_p, s)$, and $\sigma(s) \succ \sigma(t)$.

@@

\subsection{The Saturation Procedure}

>>>>

A {\em position} in a term is a tuple of natural numbers, with $\varepsilon$ denoting
the empty tuple. {\em Subterms} and positions are inductively defined as follows. Term
$t$ is a subterm of itself at position $\varepsilon$. If $s$ is a subterm of
$t_i$ at position $p$, then $s$ is a subterm of $\cst{f}(\tuplen{t})$ at
position $i.p$. A {\em context} is a term with zero or more subterms replaced by a
hole $\square$. We write $C[\overline{u}_n]$ for the term resulting from filling
in the holes of a context $C$ with the terms $\overline{u}_n$, left to right.
We say a term is {\em ground} if it has no variables. By $u[s]$
we denote that $s$ is a subterm of $u$.

@@

Applying substitution to a literal is reduced to applying it to both sides of
the (dis)equa\-tion and it is extended pointwise to clauses. It is sometimes
also written in postfix notation as $l\sigma$ and $C\sigma$. We say
$\sigmacl{C}$ is a ground instance of $C$ if $\sigma$ is a grounding
substitution.

@@

\subsection{Unification}
\label{sec:pre:unif}

Calculi from the resolution family, including superposition, perform inferences
only on unifiable terms. We say terms $s$ and $t$ are {\em unifiable} if there is a
substitution $\sigma$ such that $\sigmaterm{s} = \sigmaterm{t}$; we further say
$\sigma$ is the unifier. A \emph{unification constraint} $s \unif t$ is a pair of two terms of the same type. 
We always specify if the constraint should be interpreted as ordered or unordered.
A \emph{most general unifier} is a
unifier $\sigma$, such that for any other unifier $\theta$, there is a
substitution $\varrho$ such that $\theta = \varrho\sigma$. When there is such a
$\varrho$, we say $\sigma$ is more general than $\theta$. In first-order logic,
the most general unifier is unique up to variable renaming. Furthermore,
first-order logic admits efficient algorithms for computing the most general
unifier \cite{hv-09-unifalgs}.

In higher-order logic, unification is performed modulo rules of
$\alpha\beta\eta$ conversion. Under these conditions, the uniqueness of the most
general unifier is not guaranteed. Consider the constraint $X \,
(\cst{f} \, \cst{a}) \unif \cst{f} \, (X \, \cst{a})$. Any substitution of the
form $\{ X \mapsto \lamx{\cst{f}^i \, x} \}$ is a unifier, with $\cst{f}^i \, x$
denoting iterated, $i$-fold application of $\cst{f}$. However, neither of these substitutions
is more general than the other. To generalize most general
unifiers to higher-order logic, the concept of the {\em complete sets of unifiers} (CSU) was
introduced.

@@

\subsection{Inference Rules}
\label{sec:pre:rules}
\newcommand{\mgu}{\ensuremath{\mathit{mgu}}}

@@

\begin{tabular}{m{\dimexpr 0.36\linewidth-2\tabcolsep}m{\dimexpr 0.57\linewidth-2\tabcolsep}}
    \multicolumn{2}{l}{{\bf Equality resolution} (ER)} \\[\jot]
    $\namedinference{ER}{s \noteq t \llor C}{\sigmacl{C}}$ & where $\sigma = \mgu(s,t)$, and $\sigmacl{s \noteq t}$ is eligible for resolution
\end{tabular}

@@

\begin{tabular}{m{\dimexpr 0.36\linewidth-2\tabcolsep}m{\dimexpr 0.57\linewidth-2\tabcolsep}}
    \multicolumn{2}{l}{{\bf Equality factoring} (EF)} \\[\jot]
    $\namedinference{EF}{s \eq t \llor u \eq v \llor C}{\sigmacl{t \noteq v \llor u \eq v \llor C}}$ 
        & where $\sigma = \mgu(s,u), \sigma(t) \not\succ \sigma(s)$ and 
        $\sigmacl{s \eq t}$ is eligible for paramodulation
\end{tabular}

@@

\begin{tabular}{m{\dimexpr 0.36\linewidth-2\tabcolsep}m{\dimexpr 0.57\linewidth-2\tabcolsep}}
    \multicolumn{2}{l}{{\bf Superposition into positive literals} (SP)} \\[\jot]
    $\namedinference{SP}{s \eq t \llor C \qquad u \eq v \llor D}{\sigmacl{u[p \leftarrow t] \eq v \llor C \llor D}}$ 
        & where $\sigma = \mgu(u|_p, s), \sigma(t) \not\succ \sigma(s), \sigma(v) \not\succ \sigma(u)$,
        $\sigmacl{s \eq t}$ is eligible for paramodulation, just like $\sigmacl{u \eq v}$ and $u|_p$ is not a variable
\end{tabular}

@@

\begin{tabular}{m{\dimexpr 0.36\linewidth-2\tabcolsep}m{\dimexpr 0.57\linewidth-2\tabcolsep}}
    \multicolumn{2}{l}{{\bf Superposition into negative literals} (SN)} \\[\jot]
    $\namedinference{SN}{s \eq t \llor C \qquad u \noteq v \llor D}{\sigmacl{u[p \leftarrow t] \noteq v \llor C \llor D}}$ 
        & where $\sigma = \mgu(u|_p, s), \sigma(t) \not\succ \sigma(s), \sigma(v) \not\succ \sigma(u)$,
         $\sigmacl{s \eq t}$ is eligible for paramodulation, $\sigmacl{u \noteq v}$ is eligible for resolution, and $u|_p$ is not a variable
\end{tabular}

@@

\[
\namedsimp{RN}{s \eq t \qquad u \noteq v \llor D}
              {s \eq t \qquad \sigmacl{u[p \leftarrow \sigma(t)] \eq v \llor C \llor D}}
\]
where $\sigma = \mgu(u|_p, s)$, and $\sigma(s) \succ \sigma(t)$.

@@

\subsection{The Saturation Procedure}
\label{sec:pre:saturation}
